D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-17 ootavacuous.tex
Add Alan Stern's example invalidating (sdep U rf)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
from their relaxed atomic loads to their relaxed atomic stores, which
prevents this OOTA cycle on real-world C++ implementations.

@@

JMM Causality Test Case 12 in
Figure~\ref{lst:Causality Test Case 12}
uses arrays, which are not yet supported by the \co{herd7} tool.

@@

Figure~\ref{lst:Multi-Cycle Litmus Test}
shows an expansion to four threads of Causality Test Case 4, which is shown in
Listing~\ref{lst:Causality Test Case 4}
and discussed in
Appendix~\ref{app:Causality Test Case 4}.
The \co{herd7} tool shows multiple different possible OOTA cycles,
as expected.
However, the value returned by each thread's load directly affects the
value stored, so there is a semantic dependency in each thread.
Therefore, none of these OOTA cycles can happen in a correct real-world
C++ implementation.

@@

>>>>

Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
from their relaxed atomic loads to their relaxed atomic stores, which
prevents this OOTA cycle on correct real-world C++ implementations.

@@

JMM Causality Test Case 12 in
Listing~\ref{lst:Causality Test Case 12}
uses arrays, which are not yet supported by the \co{herd7} tool.

@@

Listing~\ref{lst:Multi-Cycle Litmus Test}
shows an expansion to four threads of Causality Test Case 4, which is shown in
Listing~\ref{lst:Causality Test Case 4}
and discussed in
Appendix~\ref{app:Causality Test Case 4}.
The \co{herd7} tool shows multiple different possible OOTA cycles,
as expected.
However, the value returned by each thread's load directly affects the
value stored, so there is a semantic dependency in each thread.
Therefore, none of these OOTA cycles can happen in a correct real-world
C++ implementation.

@@

\subsubsection{Three-Process Version of JMM Causality Test Case 4}
\label{app:Three-Process Version of JMM Causality Test Case 4}

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
shows a different way of expanding
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two processes to three processes.
The \co{herd7} tool finds OOTA values, but in a rather odd manner.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
\caption{Three-Process Version of JMM Causality Test Case 4}
\label{lst:Three-Process Version of JMM Causality Test Case 4}
\end{listing}

All processes have straightforward semantic dependencies from their
relaxed atomic loads to their relaxed atomic stores, which prevents this
OOTA cycle on correct real-world C++ implementations.

\subsubsection{Why rfe Instead of Tried-And-True rf?}
\label{app:Why rfe Instead of Tried-And-True rf?}

Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
revises
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
to flatten \co{PO()} and \co{P1()} into a single thread, so that the
resulting \co{P0()} has an rfi link instead of an rfe link connecting
the store to and load from \co{y}.
In addition, the value stored to \co{y} is forced to be at least the
value 17, and the value stored to \co{z} is forced to be at most the
value 17.
This of course means that the store to \co{z} must always be the value
17, and that a C++ implementation could prove this by examining \co{P0()}
in isolation given that it is also legal for a C++ implementation
to replace the value loaded from \co{y} with the value just stored.\footnote{
	This is not an option in C because atomic operations are volatile,
	and thus observed behavior.}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
\caption{Why rfe Instead of Tried-And-True rf?}
\label{lst:Why rfe Instead of Tried-And-True rf?}
\end{listing}

In the confines of the C++ abstract machine (omitting the no-OOTA
restriction), this allows a value-17 OOTA cycle, courtesy of the fact
that C++ respects neither control nor data dependencies.
However, the \co{herd7} tool does not identify this cycle, perhaps due
to its restrictions on its handling of arithmetic on OOTA values.

In a real-world implementation that proves that the value 17 is always
stored to \co{z}, there is no semantic dependency to
\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
atomic stores.
On the other hand, if the implementation does not prove the value
stored to \co{z}, there will be a global-time delay between each of
\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
cycle from forming.

Either way, the value-17 OOTA cycle cannot occur in a correct real-world
C++ implementation.
Yes, an implementation that proves that the value 17 is always
stored to \co{z} can satisfy the \co{exists} clause, but because
there is no semantic dependency to \co{P0()}'s store to \co{z},
this is not an OOTA cycle.

But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
of (sdep $\cup$ rfe)?

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
\caption{Why rfe Instead of Tried-And-True rf When z=17?}
\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
\end{listing}

In that case, it would be necessary to separately consider the dependency
from \co{P0()}'s load from \co{x} and \co{y} and that from its load
from \co{y} and its store to \co{z}.
The problem with this is shown in
Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
which shows an equivalent program given the knowledge that the value~17
will always be stored to \co{z}.
Because all atomic operations are relaxed and because C++ does not
respect control dependencies, the store to \co{z} can be moved up
to the very beginning of \co{P0()}.
This transformation causes the rfi link associated with \co{y} to
go backwards in time, which is perfectly legitimate in this case,
but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
problematic.

Again, in C, the load from \co{y} cannot be removed because it is
volatile, and thus observable behavior.
C++ could remove that load, but we left it in to demonstrate the all-17s
outcome.

This example demonstrates why OOTA cycles must be defined in terms
of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
\co{acyclic(sdep; rfe)}.
